---
title: "Reproducible Research: Peer Assessment 1"
author: "Fabio Correa (@feamcor, coursera, data science specialization, repdata-035)"
date: "December 2015"
output: 
  html_document:
    keep_md: true
---

```{r setup, echo=FALSE}
library(knitr)
opts_chunk$set(echo=TRUE, tidy=FALSE, cache=FALSE, fig.path='figure/')
```

## Loading the Dataset
```{r load}
filename.zip <- "activity.zip"
if(!file.exists(filename.zip)) {
    dataset.url <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
    download.file(url = dataset.url, destfile = filename.zip, method = "auto")
    message(paste(Sys.time(), "Dataset zip file downloaded!", filename.zip))
} else {
    message(paste(Sys.time(), "Dataset zip file already exists!", filename.zip))
}

filename.csv <- "activity.csv"
if(!file.exists(filename.csv)) {
    unzip(filename.zip, overwrite = TRUE)
    message(paste(Sys.time(), "Dataset file expanded!", filename.csv))
} else {
    message(paste(Sys.time(), "Dataset file already exists!", filename.csv))
}

dataset <- read.csv(filename.csv,
                    header=TRUE,
                    colClasses=c("integer", "Date", "integer"))
```
The dataset contains __`r nrow(dataset)`__ observations!  
The variables included in the dataset are:

*  __steps__: Number of steps taken in a 5-minute interval (missing values are coded as NA)
*  __date__: The date on which the measurement was taken (YYYY-MM-DD)
*  __interval__: Identifier for the 5-minute interval in which measurement was taken

## Pre-processing the Dataset
```{r preproc}
weekend <- c("Sat", "Sun")
date.type <- factor(weekdays(dataset$date,
                             abbreviate = TRUE) %in% weekend,
                    levels = c(TRUE, FALSE),
                    labels = c('weekend', 'weekday'))
dataset <- cbind(dataset, date.type)
```
New column __date.type__ added to dataset indicating whether date is __weekday__ or __weekend__.

## What is mean total number of steps taken per day?
```{r stepsday}
stepsday <- aggregate(steps ~ date, data = dataset, FUN = sum)
hist(stepsday$steps,
     freq=TRUE,
     plot=TRUE,
     main="Total Number of Steps Taken per Day",
     xlab="Number of Steps",
     ylab="Number of Days",
     col=palette("default"))
stepsday.mean <- mean(stepsday$steps)
stepsday.median <- median(stepsday$steps)
```
Individual took an average of __`r sprintf("%.0f", stepsday.mean)`__ steps per day.  
The median was a day where individual took __`r stepsday.median`__ steps.

## What is the average daily activity pattern?
## Imputing missing values
## Are there differences in activity patterns between weekdays and weekends?
